\LARGE \bf Mean Estimation from One-Bit Measurements

### 2019-12-13 proof-sgd.tex
Working on rewriting proof of theorem 2

\subsection{Proof of Theorem~\ref{thm:sgd}\eqref{item:normal-sgd}}

@@

\subsection{Proof of Theorem~\ref{thm:sgd}\eqref{item:sgd-regular}}

This proof requires somewhat more techinicality than the first part of the
theorem, including a brief detour into local asymptotic normality theory,
regular estimators, and quadratic-mean
differentiability~\cite[cf.]{VanDerVaart98}.
We assume without loss of generality that the median of the density
$f$ is 0, so that if $P_\theta$ has density $f(\cdot - \theta)$, the
median of $P_\theta$ is $\theta$.
We begin by recalling the statistical concepts we require.
\begin{definition}
  \label{definition:regular-estimator}
  A sequence of estimators $T_n$ for a parameter $\theta$ in the parametric
  family $\{P_\theta\}_{\theta \in \Theta}$ is \emph{regular at $\theta$} if
  there exists a distribution $Q$ such that for all bounded sequence
  $h_n$,
  \begin{equation*}
    \sqrt{n}(T_n - (\theta + h_n / \sqrt{n}))
    \mathop{\cd}_{P_{\theta + h_n/\sqrt{n}}}
    Q.
  \end{equation*}
\end{definition}
\begin{definition}
  \label{definition:qmd}
  Let $\{P_\theta\}_{\theta \in \Theta}$ have densities $p_\theta$
  with respect to a base measure $\mu$. The family
  is \emph{quadratic mean differentiable} (QMD) at
  $\theta$ with score $\score_\theta$ if
  \begin{equation}
    \label{eqn:qmd}
    \int \left(\sqrt{p_{\theta + h}} - \sqrt{p_\theta}
    - \half h^\top \score_\theta\sqrt{p_\theta} \right)^2 d\mu
    = o(\norm{h}^2)
  \end{equation}
  as $h \to 0$.
\end{definition}
\begin{definition}
  \label{definition:lan}
  A family of distributions $\{P_\theta\}_{\theta \in \Theta}$ is
  \emph{locally asymptotically normal with information matrix $I_\theta$} (LAN)
  at
  $\theta$ if there exists a sequence of random vectors $Z_n$ such that for
  all $h_n \to h$,
  \begin{equation*}
    \sum_{i = 1}^n \log \frac{dP_{\theta + h_n/\sqrt{n}}}{dP_\theta}
    (X_1, \ldots, X_n)
    = h^\top Z_n - \half h^\top I_\theta h + o_P(1)
  \end{equation*}
  where $Z_n \cd \normal(0, I_\theta)$ under $P_\theta$, where
  $X_i \simiid P_\theta$.
\end{definition}

@@

It remains to show that the average $\bar{\theta}_n$ is regular:
\begin{lemma}
  \label{lemma:sgd-median-regular}
  Let $h_n \to h \in \R$, and define $P_n = P_{\theta + h_n / \sqrt{n}}^n$.
  Then $\bar{\theta}_n$ is regular with
  \begin{align}
    \label{eqn:sgd-median-regular}
    \sqrt{n}\left( \bar{\theta}_n - \theta\right)
    \mathop{\cd}_{P_n}
    \normal\left( h,\frac{1}{4 f(0)^2}\right).
  \end{align}
\end{lemma}
\begin{proof}
  To show the convergence~\eqref{eqn:sgd-median-regular} we use the
  following refinement of Corollary~\ref{corollary:polyak-juditsky}, which
  provides a generalized convergence result for iteratively defined
  $\theta_n$, and whose
  proof we defer to Section~\ref{proof:normal-expansion}.
  %
  \begin{corollary}
    \label{corollary:normal-expansion}
    Let the conditions of Corollary~\ref{corollary:polyak-juditsky} hold,
    meaning that $\theta_i = \theta_{i-1} + \gamma_i \varphi(X_i -
    \theta_{i-1})$ for $X_i = \theta + Z_i$, where $Z_i$ are i.i.d.\ with
    $\E[\varphi(Z)] = 0$. Additionally assume the local smoothness
    condition that there exist $0 < \lambda \le 1$ and $K < \infty$ such
    that
    \begin{equation}
      \label{eqn:additional-local-smooth}
      \E[|\varphi(x + Z) - \varphi(Z)|^2]
      \le K (|x|^\lambda + x^2).
    \end{equation}
    Set $\Delta_i = \theta_i - \theta$ and $\bar{\Delta}_i = \frac{1}{n}
    \sum_{i=1}^n \Delta_i$. Then
    \begin{enumerate}[(i)]
    \item \label{item:regularity}
      The sequence $\Delta_i$ is regular, that is,
      \begin{equation}
        \sqrt{n} \bar{\Delta}_n
        = -\frac{1}{\sqrt{n}} \frac{1}{\psi'(0)} \sum_{i=1}^{n-1} \varphi(Z_i)
        + o_{P,n}(1).
        \label{eq:normal_expansion_lem}
      \end{equation}
    \item \label{item:apply-le-cam} Let $Z_i$ as in
      Corollary~\ref{corollary:polyak-juditsky} have absolutely continuous
      density $p$ with median $0$, define $\score_h(z) = \frac{p'(z - h)}{p(z
        - h)}$, and assume that $I_h \defeq \E_\theta[\score_h(Z)^2]$ is
      continuous in $h$ near 0.  Then for any converging sequence $h_n \to h$,
      \begin{equation*}
        \sqrt{n} \bar{\Delta}_n
        \mathop{\cd}_{P_{\theta + h_n/\sqrt{n}}^n}
        \normal\left( \frac{-h}{\psi'(0)} \E_p[\varphi(Z) \score_0(Z)],
        \frac{\chi(0)}{\psi'^2(0)} \right).
      \end{equation*}
    \end{enumerate}
  \end{corollary}

@@

\subsubsection*{Proof of Theorem~\ref{thm:sgd}\eqref{item:sgd-ms-convergence}}

We begin with the following result from \cite{polyak1990new}:
\begin{corollary}[\cite{polyak1990new}, Theorem 2]
  \label{corollary:polyak-mse}
  Let
  \begin{align} \label{eq:polyak_new_measurements}
    \begin{cases}
      U_n = U_{n-1} - \gamma_n \varphi(Y_n), & Y_n = g'(U_{n-1})+Z_n \\
      \bar{U}_n= \frac{1}{n} \sum_{i=1}^n U_n, & n=1,2,\ldots.
    \end{cases}
  \end{align}
  Assume that the function $g(x)$ is twice differentiable with a strictly
  positive and uniformly bounded second derivative. In particular, $g(x)$ is
  convex with a unique minimizer $x^\star \in \R$. Moreover, assume that the
  noises $Z_n$ are uncorrelated and identically distributed with a
  distribution with a density for which the Fisher information exits. Let
  $\psi(x)$ and $\chi(x)$ be defined as in
  Theorem~\ref{corollary:polyak-juditsky}-(iii) and satisfy the conditions
  there. Assume in addition that $\chi(0)>0$, condition
  \eqref{eqn:local-hessian-psi} with $\lambda = 1$, and there exits $K_3$ such
  that
  \begin{equation*}
    \ex{  | \varphi(x+Z_1) |^4 } \leq K_3(1+|x|^4). 
  \end{equation*}
  Finally, assume that the sequence $\{\gamma_n \}$ satisfies conditions \eqref{eq:conditions1} and \eqref{eq:conditions2}. Then
  \begin{equation*}
    V_n \triangleq \ex{ \left(\bar{U}_n-x^\star \right)^2 } = n^{-1}\frac{\chi(0)} { (\psi'(0))^2 (g''(x^\star))^2 } + o(n^{-1}).
  \end{equation*}
\end{corollary}

We now use Theorem~\ref{thm:polyak_new} with $g(x) = 0.5(x-\theta)^2$, $\varphi(x) = \sgn(x)$, $Z_n = \theta-X_n$. From \eqref{eq:polyak_new_measurements} we have
\begin{align*} 
U_n & = U_{n-1} + \gamma_n \sgn(\theta-U_{n-1} - Z_n )  \\
& = U_{n-1} + \gamma_n \sgn(X_n-U_{n-1} ),
\end{align*}
so the estimator $\bar{U}_n$ equals to the one defined by \eqref{eq:sgd_est} and \eqref{eq:sgd_alg}. Note that
\begin{equation*}
\ex{ | \varphi(x+Z_1) |^4 } = 1 \leq K_3(1+|x|^4)
\end{equation*}
for any $K_3\geq 1$, the Fisher information of $Z_1$ is $\sigma^2$, $\chi(x) = 1 > 0$, and that 
the conditions in Theorem~\ref{thm:polyak_new} on $\psi(x)$ and $\chi(x)$ were verified to hold in the first part of the proof. In particular, $\psi'(0) = (2f(0))^{-2}$. Since $f(x)$ satisfies the conditions above with $x^\star = \theta$ and $g''(x) = 1$. Theorem~\ref{thm:polyak_new} implies that for any $\theta \in \R$, 
\begin{equation*}
V_n = \ex{ \left({\theta}_n-\theta \right)^2 } = \frac{1}{4n f^2(0)} + o(n^{-1}).
\end{equation*}

>>>>

\subsection{Proof of Theorem~\ref{thm:sgd}\eqref{item:normal-sgd}}
\label{sec:proof-normal-sgd}

@@

\subsection{Proof of Theorem~\ref{thm:sgd}\eqref{item:sgd-regular}}
\label{sec:proof-sgd-regular}

This proof requires somewhat more technicality than the first part of the
theorem, including a brief detour into local asymptotic normality theory,
regular estimators, and quadratic-mean
differentiability~\cite[cf.]{VanDerVaart98}.
We assume without loss of generality that the median of the density
$f$ is 0, so that if $P_\theta$ has density $f(\cdot - \theta)$, the
median of $P_\theta$ is $\theta$.
We begin by recalling the statistical concepts we require.
\begin{definition}
  \label{definition:regular-estimator}
  A sequence of estimators $T_n$ for a parameter $\theta$ in the parametric
  family $\{P_\theta\}_{\theta \in \Theta}$ is \emph{regular at $\theta$} if
  there exists a distribution $Q$ such that for all bounded sequence
  $h_n$,
  \begin{equation*}
    \sqrt{n}(T_n - (\theta + h_n / \sqrt{n}))
    \mathop{\cd}_{P_{\theta + h_n/\sqrt{n}}}
    Q.
  \end{equation*}
\end{definition}
\begin{definition}
  \label{definition:qmd}
  Let $\{P_\theta\}_{\theta \in \Theta}$ have densities $p_\theta$
  with respect to a base measure $\mu$. The family
  is \emph{quadratic mean differentiable} (QMD) at
  $\theta$ with score $\score_\theta$ if
  \begin{equation}
    \label{eqn:qmd}
    \int \left(\sqrt{p_{\theta + h}} - \sqrt{p_\theta}
    - \half h^\top \score_\theta\sqrt{p_\theta} \right)^2 d\mu
    = o(\norm{h}^2)
  \end{equation}
  as $h \to 0$.
\end{definition}
\begin{definition}
  \label{definition:lan}
  A family of distributions $\{P_\theta\}_{\theta \in \Theta}$ is
  \emph{locally asymptotically normal with information matrix $I_\theta$} (LAN)
  at
  $\theta$ if there exists a sequence of random vectors $Z_n$ such that for
  all $h_n \to h$,
  \begin{equation*}
    \sum_{i = 1}^n \log \frac{dP_{\theta + h_n/\sqrt{n}}}{dP_\theta}
    (X_1, \ldots, X_n)
    = h^\top Z_n - \half h^\top I_\theta h + o_P(1)
  \end{equation*}
  where $Z_n \cd \normal(0, I_\theta)$ under $P_\theta$, where
  $X_i \simiid P_\theta$.
\end{definition}

@@

It remains to show that the average $\bar{\theta}_n$ is regular:
\begin{lemma}
  \label{lemma:sgd-median-regular}
  Let $h_n \to h \in \R$, and define $P_n = P_{\theta + h_n / \sqrt{n}}^n$.
  Then $\bar{\theta}_n$ is regular with
  \begin{align}
    \label{eqn:sgd-median-regular}
    \sqrt{n}\left( \bar{\theta}_n - \theta - h_n / \sqrt{n}\right)
    \mathop{\cd}_{P_n}
    \normal\left( h,\frac{1}{4 f(0)^2}\right).
  \end{align}
\end{lemma}
\begin{proof}
  To show the convergence~\eqref{eqn:sgd-median-regular} we use the
  following refinement of Corollary~\ref{corollary:polyak-juditsky}, which
  provides a generalized convergence result for iteratively defined
  $\theta_n$, and whose
  proof we defer to Section~\ref{proof:normal-expansion}.
  %
  \begin{corollary}
    \label{corollary:normal-expansion}
    Let the conditions of Corollary~\ref{corollary:polyak-juditsky} hold,
    meaning that $\theta_i = \theta_{i-1} + \gamma_i \varphi(X_i -
    \theta_{i-1})$ for $X_i = \theta + Z_i$, where $Z_i$ are i.i.d.\ with
    $\E[\varphi(Z)] = 0$. Additionally assume the local smoothness
    condition that there exist $0 < \lambda \le 1$ and $K < \infty$ such
    that
    \begin{equation}
      \label{eqn:additional-local-smooth}
      \E[|\varphi(x + Z) - \varphi(Z)|^2]
      \le K (|x|^\lambda + x^2).
    \end{equation}
    Set $\Delta_i = \theta_i - \theta$ and $\bar{\Delta}_i = \frac{1}{n}
    \sum_{i=1}^n \Delta_i$. Then
    \begin{enumerate}[(i)]
    \item \label{item:regularity}
      The sequence $\Delta_i$ is regular, that is,
      \begin{equation}
        \sqrt{n} \bar{\Delta}_n
        = -\frac{1}{\sqrt{n}} \frac{1}{\psi'(0)} \sum_{i=1}^{n-1} \varphi(Z_i)
        + o_{P,n}(1).
        \label{eq:normal_expansion_lem}
      \end{equation}
    \item \label{item:apply-le-cam} Let $Z_i$ as in
      Corollary~\ref{corollary:polyak-juditsky} have absolutely continuous
      density $p$ with median $0$, define $\score_h(z) = \frac{p'(z - h)}{p(z
        - h)}$, and assume that $I_h \defeq \E_\theta[\score_h(Z)^2]$ is
      continuous in $h$ near 0.  Then for any converging sequence $h_n \to h$,
      \begin{equation*}
        \sqrt{n} \bar{\Delta}_n
        \mathop{\cd}_{P_{\theta + h_n/\sqrt{n}}^n}
        \normal\left( \frac{-h}{\psi'(0)} \E_p[\varphi(Z) \score_0(Z)],
        \frac{\chi(0)}{\psi'^2(0)} \right).
      \end{equation*}
    \end{enumerate}
  \end{corollary}

@@

\subsection{Proof of Theorem~\ref{thm:sgd}\eqref{item:sgd-ms-convergence}}
\label{sec:proof-sgd-ms-convergence}

We begin with the following result from \cite{polyak1990new}:
\begin{corollary}[\cite{polyak1990new}, Theorem 2]
  \label{corollary:polyak-mse}
  Define the iteration
  \begin{align} \label{eq:polyak_new_measurements}
    \begin{cases}
      U_n = U_{n-1} - \gamma_n \varphi(Y_n), & Y_n = g'(U_{n-1})+Z_n \\
      \bar{U}_n= \frac{1}{n} \sum_{i=1}^n U_n, & n=1,2,\ldots.
    \end{cases}
  \end{align}
  Assume that the function $g$ is $\mc{C}^2$, strictly convex, has Lipschitz
  derivative, and is minimized by $x^\star$. Moreover, assume that the
  noises $Z_n$ are i.i.d.\ with density $p$ and
  that the Fisher information $\E[(p'(Z))^2 / p(Z)^2]$ exists and is finite.
  Let $\psi(x)$ and $\chi(x)$ be defined as in
  Corollary~\ref{corollary:polyak-juditsky} and satisfy the conditions in
  the corollary. Assume in addition that $\chi(0)>0$, condition
  \eqref{eqn:local-hessian-psi} with $\lambda = 1$, and there exits $K_3$
  such that
  \begin{equation*}
    \ex{  | \varphi(x+Z_1) |^4 } \leq K_3(1+|x|^4). 
  \end{equation*}
  Finally, assume that the sequence $\{\gamma_n \}$ satisfies conditions
  \eqref{eqn:lazy-gamma} and \eqref{eqn:stringent-gamma}. Then
  \begin{equation*}
    V_n \defeq
    \E\Big[(\bar{U}_n-x^\star )^2\Big] = n^{-1}\frac{\chi(0)} {
      (\psi'(0))^2 (g''(x^\star))^2 } + o(n^{-1}).
  \end{equation*}
\end{corollary}

We apply Corollary~\ref{corollary:polyak-mse} with $g(x) = 0.5(x-\theta)^2$,
$\varphi(x) = \sgn(x)$, $Z_n = \theta-X_n$.
The update~\eqref{eq:polyak_new_measurements} gives
\begin{align*} 
  U_n %% & = U_{n-1} + \gamma_n \sgn(\theta-U_{n-1} - Z_n )  \\
  & = U_{n-1} + \gamma_n \sgn(X_n-U_{n-1} ),
\end{align*}
so the estimator $\bar{U}_n$ is identical to the stochastic gradient
estimator~\eqref{eq:sgd_est}. We have $\E[\varphi(x + Z)^4] = 1$ and by
assumption the Fisher information $\E[(f'(Z))^2 / f(Z)^2]$ exists, and the
functions $\psi$ and $\chi$ have the desired conditions of
Corollary~\ref{corollary:polyak-juditsky} (as we verify in
Section~\ref{sec:proof-normal-sgd}). Finally, the function $\theta \mapsto
\E_{P_\theta}[(\bar{\theta}_n - \theta)^2]$ is continuous in $\theta$, so
that for $x^\star = \theta$ and $g'' \equiv 1$, we may apply
Corollary~\ref{corollary:polyak-mse} and a finite approximation to the prior
measure $\pi$ to obtain
Theorem~\ref{thm:sgd}\eqref{item:sgd-ms-convergence}.

%% \begin{equation*}
%% \ex{ | \varphi(x+Z_1) |^4 } = 1 \leq K_3(1+|x|^4)
%% \end{equation*}
%% for any $K_3\geq 1$, the Fisher information of $Z_1$ is $\sigma^2$, $\chi(x) = 1 > 0$, and that 
%% the conditions in Theorem~\ref{thm:polyak_new} on $\psi(x)$ and $\chi(x)$ were verified to hold in the first part of the proof. In particular, $\psi'(0) = (2f(0))^{-2}$. Since $f(x)$ satisfies the conditions above with $x^\star = \theta$ and $g''(x) = 1$. Theorem~\ref{thm:polyak_new} implies that for any $\theta \in \R$, 
%% \begin{equation*}
%% V_n = \ex{ \left({\theta}_n-\theta \right)^2 } = \frac{1}{4n f^2(0)} + o(n^{-1}).
%% \end{equation*}
